[Re] Spike Synchronization and Rate Modulation Differentially Involved in Motor Cortical Function

### 2017-03-29 article/VRostami_JIto_MDenker_SGruen_2016.tex
changes after reviewrs comments

Understanding how information is processed by networks of neurons in the
brain is a major goal in neuroscience. There has been a long-standing
debate in the community on the contribution of the firing activity of
individual neurons or populations of neurons to information processing:
one perspective is that the firing rate of the neurons, i.e.~the number
of spikes per time unit, represents information or is relevant for the
processing. Another perspective is that neurons communicate through
precisely timed coordination of spiking, a view that results from the
insight that a neuron fires most efficiently if it receives synchronous
spike input, i.e., in the range of a few ms \autocite{Abeles82}.

To test whether temporal coordination of spiking activity is indeed
relevant for neuronal information processing, advanced data analysis
methods are required that perform correlation analysis between
simultaneously recorded single unit spike trains. The Unitary Events
(UE) analysis method \autocites{GruenPhD}{Gruen99}{Gruen02a}{Gruen02b}
is able to extract significant spike synchrony between neuronal
activities that is beyond what is expected by chance (given the rates)
and to follow the dynamics of this coordination. The tricky issue with
such analyses is that one has to take into account that experimental
spike data are typically non-stationary in the sense that the firing
rates are not stationary in time and not homogeneous across trials, and
that the statistics of the individual spike trains deviate from those of
a Poisson process \autocites{Gruen09}{GruenRotter10_Chap10}. If such
features are ignored there is a considerable danger of occurrence of
false positives and thus wrong interpretation of the data. In
particular, changes in the firing rates are the most prominent
generators of false positives if ignored. The original UE method
considers this aspect by performing the analysis in a sliding time
window fashion. In later versions of the analysis method also other
features were corrected for by considering the respective features in
the null-hypothesis of the tests
\autocites{Gruen03b}{Maldonado08}{Louis10}{Pipa2013}, either by using
extended analytical descriptions of the null-hypotheses or by use of
surrogate methods \autocites{Gruen09}{GruenRotter10_Chap10}{Louis10}.

@@

Several of these studies based on the UE method, in particular
\autocite{Riehle97}, have been widely cited and the method has been
recognized as one of the standard tools to analyze temporal coordination
of neuronal spiking activities \autocites{Brown04_456}{Nakahara02}. The
analysis method has been and is taught in international data analysis
courses. However, a publicly available and open source implementation of
the UE method had not been available. Only very recently, one of the
authors of this study (VR) reimplemented the UE analysis as part of the
Electrophysiology Analysis Toolbox\footnote{\url{http://neuralensemble.org/elephant/}}
(Elephant), a Python library that provides implementations for the
analysis of electrophysiological data. A reimplementation of the method,
rather than releasing the existing code, was justified by two factors.

@@

In the following, we will explain the algorithm in detail. The primary
data entering the UE method is a set of parallel, i.e., simultaneously
recorded, spike trains, recorded in one or multiple trials. In Elephant,
an individual spike train of a particular neuron in a particular trial
data is represented as a \texttt{Spiketrain} object in the data object
model provided by the Neo library, which stores the time points of spike
occurrences along with additional information describing the spike
train, such as the start and end times. The UE method consumes a nested
list of \texttt{Spiketrain} objects relating to the spike trains of
individual neurons in the individual trials. In order to perform the
necessary calculations, the spike data must be converted to an
alternative time-binned representation (line 1 in the pseudo code) where
the parallel spike trains are stored as binary sequences of ones
(marking time bins containing at least 1 spike) and zeros (bins with no
spike). The bin size is a parameter provided as input (parameter
\emph{\texttt{bin\_size}} in the pseudo code) to the analysis, and
defines the temporal precision in detecting spike synchrony. A spike
pattern, which is a representation of spike synchrony in the analysis,
is defined as a specific vector of zeros and ones in one time bin of a
given trial across all neurons. Thus, the number of possible spike
patterns is given by \(2^{N}-N-1\), where \(N\) is the number of
neurons. In order to limit the analysis to a set of patterns of
interest, the input parameter \emph{\texttt{pattern\_hash\_values}}
specifies the patterns to consider in the analysis in the form of a hash
value which uniquely represents each spike pattern (line 2 in the pseudo
code). The hash value is obtained by interpreting the binary spike
pattern as a binary number, where the \(n\)-th neuron is represented by
bit \(n-1\).

@@

In case of selecting \emph{``analytic\_TrialAverage''} - as used in the
remainder of this study for the reproduction of \autocite{Riehle97} -
the number of spikes per neuron within the sliding window is summed
across trials and divided by the number of trials and bins contained,
thus yielding the average probability \(p_{i}\) to have a spike of
neuron \(i\) in a bin of the time window. The probability to find a
particular pattern by chance in a bin is then computed by multiplication
of the relevant probabilities, e.g.~for a pattern {[}1,0,1{]} the
probability \(p\) of occurrence is given by
\(p_{[1,0,1]}=p_{1}*(1-p_{2})*p_{3}\). Note that (\(1-p_{2}\)) is the
probability for neuron 2 to contribute no spike to the pattern. The
expected number of pattern occurrences, computed as the product of the
occurrence probability \(p\) of the pattern, e.g. \(p_{[1,0,1]}\), and
the number of bins (across all trials) covered by the sliding window.
The distribution of pattern occurence numbers is given by a Poisson
distribution with the mean equal to the expected number.

Alternatively, for the \emph{``analytic\_TrialByTrial''} method, the
firing probability of each neuron is calculated in a trial-by-trial
manner based on the spike counts per trial. The probability \(p\) of
finding the pattern by chance is calculated by summing the products of
the firing probabilities obtained individually from each trial. As for
the trial-averaging method, the expected number of pattern occurrences
is given by multiplication with the number of bins of the trial in the
sliding window, used as the mean of a Poisson distribution to obtain the
distribution of expected pattern occurence numbers.

@@

Our reimplementation uses the \texttt{conversion} module of Elephant for
converting the spike data (represented as a series of time stamps) into
the binary sequence to guarantee a unique, global binning mechanism for
all analysis methods provided in Elephant. The bin size to be set for
the analysis was extracted from the original publication. However,
defining the time point to start the binning of each single trial data
required to know the alignment event in each trial and how much time
before this event (pre-time) is considered. Since this information was
not documented in the original paper, we tried several possibilities
until we got an agreement with the original figures as will be shown in
the Results.

To check if our Python implementation produces the same results as the
implementation(s) used in the original publication, we compare each of
our figures visually in detail with the original figures. This is also
the test if we used the correct data files since the names of the data
sets are not mentioned in the original publication. A direct numerical
comparison to the original results plotted in these figures was not
possible as this data is not available.

@@

For the reproduction of the original results in \autocite{Riehle97}, we
focus on reproducing Figure 2A-F and Figure 4A. Figure 2 is chosen
because it represents the main result of the study and includes the UE
analysis that underlies all subsequent analyses. Figure 4A is chosen
because this is an example of the application of the UE method to data
with more than 2 neurons. In terms of complexity of the code the
implementation of the UE analysis for three or more neurons is
considerably more demanding than for only two neurons. With this example
we show that our implementation is capable of performing the UE analysis
for the generic case of arbitrary numbers of neurons.

@@

\begin{figure}[htbp]
\centering
\includegraphics{figure1.eps}
\caption{\label{fig:figure2alignedPS}Initial attempt to reproduce Figure
2 of the original publication with trial alignment to PS. \textbf{A)}
Raster plot of two neurons (neuron 2: top of panel; neuron 3: bottom of
panel) in 32 trials (sorted identically for both neurons). \textbf{B)}
Average firing rate of each neuron calculated across trials in a sliding
window of length 100~ms in steps of 5~ms. \textbf{C)} Same raster plot
as in panel A with spike coincidences (i.e., pattern {[}1,1{]}) between
the two neurons marked by cyan squares. \textbf{D)} Empirical (cyan) and
expected (magenta) number of coincidences calculated in a time-resolved
manner (parameters of sliding window identical to panel B). \textbf{E)}
Time course of the surprise measure, calculated in same sliding windows
as in panel B. Surprise values that correspond to positive and negative
significance levels \(\alpha+=0.05\) and \(\alpha-=0.95\) are shown with
by horizontal red and green lines, respectively. \textbf{F)} Same raster
plot as in panel A with significant coincidences, i.e.~UEs, marked by
red squares.}\label{fig:figure2alignedPS}
\end{figure}

@@

The original publication does not provide information which event was
used as the trigger. In this experiment, 2 events that occur in every
trial could serve as trigger events, the preparatory signal PS (event
code 114) and the response signal RS (event code 124). We noticed that
the time interval between PS and RS for the longest trials was not
identical across the respective trials and varied by \(\pm\) 1~ms. Given
the UE method is applied on a time scale of 5~ms, the analysis results
therefore are expected to depend on whether trials are aligned to PS or
RS. Thus, we decide to generate the results for both alignments.

Figures~\ref{fig:figure2alignedPS} and ~\ref{fig:figure2alignedRS} show
the results of performing the UE analysis for PS- and RS-aligned data,
respectively. Here, the analysis parameters are set to the identical
values as reported in the original publication (bin size: 5~ms, analysis
time window size: 100~ms, time step of the sliding window: 5~ms,
significance level \(\alpha= 0.05\)). The comparison of the two figures
to the original figure shows agreement in the raster displays (panel A)
and the time-resolved, trial-averaged firing rate estimates (B).
However, although the graphs of the number of coincidences per sliding
window (panel D) and the surprise measure (panel E) are similar in their
overall general behavior, they differ in the details. Thus, indeed the
choice of the alignment influences the analysis result. In order to test
if one of the two alignments is in agreement with the original
publication, we perform a detailed visual comparison of our two figures
and the original one on the basis of the spikes marked as coincident
(panels C) and as part of a UE (panel D). We notice that when aligning
to PS (Figure~\ref{fig:figure2alignedPS}), the marked spikes do not
agree in all details with the original figure. However,
Figure~\ref{fig:figure2alignedRS}, with trials aligned to RS, agrees
completely with the original figure. Although we cannot check the
identity to a further extent because the exact numerical values of the
original results are not available anymore in electronic form, we
believe that this visual inspection provides enough support for a
reproduction of the original result by the reimplementation of the UE
method.

@@

\begin{figure}[htbp]
\centering
\includegraphics{PS_aligned_matlabdata_marked.eps}
\caption{\label{fig:comparison_raster}Close-ups of the original raster
display in Figure 4A of the original publication (left) and the first
data file available at hand for reproduction (right) reveal slight
differences in the positions of some spikes. Data on the right are
aligned (similar to the data on the left) to ES1 (event code 15 in the
GDF data file). The time before the alignment event is chosen as 700~ms,
and a bin size of 5~ms is used. The red marks indicate spike clusters
with identified differences between the left and the right panel, where
at least one spike is shifted in the right panel compared to the left
one.}\label{fig:comparisonux5fraster}
\end{figure}

@@

In our reproduction of Figure 2 of the original paper we use
preprocessed data available in 1 ms resolution, that likely experienced
the \texttt{alexa2gdf} program for conversion as shown in
Figure~\ref{fig:workflow_report} (the rightmost branch), before data are
loaded into our reimplementation of the UE analysis. However, according
to the aforementioned report by MD, we only have a chance to reproduce
Figure 4A of \autocite{Riehle97} if we have the original data or a
version of them with a time resolution lower than 1 ms available. The
original raw data with 0.1~ms resolution are presumably only available
on a storage medium and format that at present we are not able to read
and interpret. However, after we contacted AR she found the data
(``jenny201\_345\_preprocessed.gdf'') of Figure 4A of
\autocite{Riehle97} with a time resolution of 0.5~ms, which likely
experienced the \texttt{2gdf} program for conversion (middle left box in
Figure~\ref{fig:workflow_report}).

We loaded this data at 0.5~ms resolution into Python and converted the
data from the 0.5~ms to the required 1~ms resolution by the mathematical
operation \(\left\lfloor x+\frac{1}{2}\right\rfloor\) , called
``rounding half up''. In numerical software packages, including IDL,
this operation is typically implemented by a function named round().
However, the round() implementation of NumPy (version 1.11.0) performs
an even rounding, i.e., values exactly halfway between two integers are
rounded to the nearest even integer. Indeed, the latter implementation
of rounding did not reproduce the result of the original publication.
Thus, we used the expression \texttt{floor(x+0.5)} to perform rounding
as it is implemented by IDL. The procedure completely reproduces panel A
of Figure 4 in the original publication (see
Figure~\ref{fig:reproducedFig4A}).

@@

We are able to reproduce the original results of \autocite{Riehle97} by
applying a new reimplementation of the Unitary Events analysis method in
Python to the original data. The method involves a number of numerical
computations and is very sensitive, as we show here by the comparison of
Figures ~\ref{fig:figure2alignedPS} and ~\ref{fig:figure2alignedRS},
which differed in the events that the trials were aligned to. This
difference in the alignment would not make any difference in the results
if the time difference between the two events (PS and RS) were identical
across the trials. But since the latter was not the case due to hardware
features of the recording setup (as we learned from the first author of
the original publication), the binning of the data started at a slightly
different time points in different trials. This likely led to a loss or
an addition of a spike in a bin and thus to a small difference of the
number of spike synchrony events (see also the discussion on the issues
of exclusive binning in \autocite{Gruen99}). Nevertheless, this
sensitivity of the method is a strong indication that our reproductions
of Figure 2 and Figure 4A of the original publication are correct and
that our new implementation of the analysis faithfully implemented the
UE method.

@@

The reproduction of Figure 4A is more complicated than reproducing
Figure 2 due to additional reasons. First of all, the data were not
available to us. After requesting them from the first author of the
original publication we received data and were not able to reproduce the
result - in terms of the UE results, but also the data seemed slightly
different. After further consultation with the original authors we
learned that the original Figure 4A was not generated by the Matlab
implementation used for Figure 2 but by another implementation in IDL.
Both are not available to us. However, we were told that the third
author of the original publication performed a thorough comparison of
the two implementations at the time and the final report on that
investigation was made available to us. This enables us to define the
correct workflow that reproduces the original result, given we have the
data in the correct resolution at hand.

@@

\section{Acknowldegements}\label{acknowldegements}

This project received funding from the European Union's Horizon 2020
research and innovation programme under grant agreement No. 720270,
Deutsche Forschungsgemeinschaft Grant DE 2175/2-1 and GR 1753/4-2 of the
Priority Program (SPP 1665), the German-Japanese Computational
Neuroscience Project (German Federal Ministry for Education and
Research, BMBF Grant 01GQ1114), from the Helmholtz Portfolio Theme
``Supercomputing and Modeling for the Human Brain'', and from the Osaka
Univ for the project `Neural mechanism of active vision studied by
combining large-scale sampling of neural activity and advanced
computational analysis'.

>>>>

Understanding how information is processed by networks of neurons in the
brain is a major goal in neuroscience. There has been a long-standing
debate in the community on the contribution of the firing activity of
individual neurons or populations of neurons to information processing:
one perspective is that the firing rate of the neurons, i.e.~the number
of spikes per time unit, represents information or is relevant for the
processing. Another perspective is that neurons communicate through
precisely timed coordination of spiking, a view that results from the
insight that a neuron fires most efficiently if it receives synchronous
spike input, i.e., in the range of a few milliseconds
\autocite{Abeles82}.

To test whether temporal coordination of spiking activity is indeed
relevant for neuronal information processing, advanced data analysis
methods are required that perform correlation analysis between
simultaneously recorded single unit spike trains. The Unitary Events
(UE) analysis method \autocites{GruenPhD}{Gruen99}{Gruen02a}{Gruen02b}
is able to extract significant spike synchrony between neuronal
activities that is beyond what is expected by chance (given the rates)
and to follow the dynamics of this coordination. The tricky issue with
such analyses is that one has to take into account that experimental
spike data are typically non-stationary in the sense that the firing
rates are not stationary in time and not homogeneous across trials, and
that the statistics of the individual spike trains deviate from those of
a Poisson process \autocites{Gruen09}{GruenRotter10_Chap10}. If such
features are ignored there is a considerable danger of occurrence of
false positives and thus wrong interpretation of the data. In
particular, changes in the firing rates are the most prominent
generators of false positives if ignored. The original UE method
considers this aspect by performing the analysis in a sliding time
window fashion. In later versions of the analysis method other features
were corrected for by considering the respective features in the
null-hypothesis of the tests
\autocites{Gruen03b}{Maldonado08}{Louis10}{Pipa2013}, either by using
extended analytical descriptions of the null-hypotheses or by use of
surrogate methods \autocites{Gruen09}{GruenRotter10_Chap10}{Louis10}.

@@

Several of these studies based on the UE method, in particular
\autocite{Riehle97}, have been widely cited and the method has been
recognized as one of the standard tools to analyze temporal coordination
of neuronal spiking activities \autocites{Brown04_456}{Nakahara02}. The
analysis method has been and is taught in international data analysis
courses. However, a publicly available and open source implementation of
the UE method had not been available. Only very recently, one of the
authors of this study (VR) reimplemented the UE analysis as part of the
Electrophysiology Analysis Toolbox\footnote{\url{http://neuralensemble.org/elephant/}}
(Elephant), a Python library that provides implementations for the
analysis of electrophysiological data. This reimplementation of the
method was particularly required for, in addition to the general
motivation for a reimplementation to confirm the results not depending
on small implementation details or mistakes, the following two specific
reasons.

@@

In the following, we will explain the algorithm in detail. The primary
data entering the UE method is a set of parallel, i.e., simultaneously
recorded, spike trains, recorded in one or multiple trials. In Elephant,
an individual spike train of a particular neuron in a particular trial
data is represented as a \texttt{Spiketrain} object in the data object
model provided by the Neo library, which stores the time points of spike
occurrences along with additional information describing the spike
train, such as the start and end times. The UE method consumes a nested
list of \texttt{Spiketrain} objects relating to the spike trains of
individual neurons in the individual trials. In order to perform the
necessary calculations, the spike data must be converted to an
alternative time-binned representation (line 1 in the pseudo code) where
the parallel spike trains are stored as binary sequences of ones
(marking time bins containing at least 1 spike) and zeros (bins with no
spike). The bin size is a parameter provided as input (parameter
\emph{\texttt{bin\_size}} in the pseudo code) to the analysis, and
defines the temporal precision in detecting spike synchrony. A spike
pattern, which is a representation of spike synchrony in the analysis,
is defined as a specific vector of zeros and ones in one time bin of a
given trial across all neurons. Since the aim of the UE method is to
detect significant spike synchrony, the total number of spike patterns
of concern is given by \(2^{N}-N-1\), where \(N\) is the number of
neurons, the first term is the number of all possible spike patterns,
the second term is the number of spike patterns composed of only one
spike, and the third term is the number of the pattern without a spike.
In order to limit the analysis to a set of patterns of interest, the
input parameter \emph{\texttt{pattern\_hash\_values}} specifies the
patterns to consider in the analysis in the form of a hash value which
uniquely represents each spike pattern (line 2 in the pseudo code). The
hash value is obtained by interpreting the binary spike pattern as a
binary number, where the \(n\)-th neuron is represented by bit \(n-1\).

@@

In case of selecting \emph{``analytic\_TrialAverage''} - as used in the
remainder of this study for the reproduction of \autocite{Riehle97} -
the number of spikes per neuron within the sliding window is summed
across trials and divided by the number of trials and bins contained,
thus yielding the average probability \(p_{i}\) to have a spike of
neuron \(i\) in a bin of the time window. The probability to find a
particular pattern by chance in a bin is then computed by multiplication
of the relevant probabilities, e.g.~for a pattern {[}1,0,1{]} the
probability \(p\) of occurrence is given by
\(p_{[1,0,1]}=p_{1}*(1-p_{2})*p_{3}\). Note that (\(1-p_{2}\)) is the
probability for neuron 2 to contribute no spike to the pattern. The
expected number of pattern occurrences, computed as the product of the
occurrence probability \(p\) of the pattern, e.g. \(p_{[1,0,1]}\), and
the number of bins (across all trials) covered by the sliding window.
The distribution of pattern occurrence numbers is given by a Poisson
distribution with the mean equal to the expected number.

Alternatively, for the \emph{``analytic\_TrialByTrial''} method, the
firing probability of each neuron is calculated in a trial-by-trial
manner based on the spike counts per trial. The probability \(p\) of
finding the pattern by chance is calculated by summing the products of
the firing probabilities obtained individually from each trial. As for
the trial-averaging method, the expected number of pattern occurrences
is given by multiplication with the number of bins of the trial in the
sliding window, used as the mean of a Poisson distribution to obtain the
distribution of expected pattern occurrence numbers.

@@

Our reimplementation uses the \texttt{conversion} module of Elephant for
converting the spike data (represented as a series of timestamps) into
the binary sequence to guarantee a unique, global binning mechanism for
all analysis methods provided in Elephant. The bin size to be set for
the analysis was extracted from the original publication. However,
defining the time point to start the binning of each single trial data
required to know the alignment event in each trial and how much time
before this event (pre-time) is considered. Since this information was
not documented in the original paper, we tried several possibilities
until we got an agreement with the original figures as will be shown in
the Results.

To check if our Python implementation produces the same results as the
implementation(s) used in the original publication, we compare each of
our figures in detail with the original figures. For this comparison, as
the original results used to generate these figures are not available to
us, we first extract the times of the spikes and unitary events from the
vector graphic image (PDF) in the original paper. Then, we directly
compare these times to the times of spikes and unitary events in our
reproduced results by plotting the former against the latter, as well as
examining the distribution of the differences between them.

@@

For the reproduction of the original results in \autocite{Riehle97}, we
have to focus on reproducing Figure 2A-F and Figure 4A, since for the
other figures the data still available were either incomplete in respect
to metadata (original Figure 3) or all the required data set were not
available anymore (original Figure 4B,C) . Figure 2 represents the main
result of the study and includes the UE analysis that underlies the
subsequent analyses. Figure 4A is an example of the application of the
UE method to data with more than 2 neurons. In terms of complexity of
the code the implementation of the UE analysis for three or more neurons
is considerably more demanding than for only two neurons. With this
example we show that our implementation is capable of performing the UE
analysis for the generic case of arbitrary numbers of neurons.

@@

\begin{figure}[htbp]
\centering
\includegraphics{figure1.eps}
\caption{\label{fig:figure2alignedPS}Initial attempt to reproduce Figure
2 of the original publication with trial alignment to PS. \textbf{A)}
Raster plot of two neurons (neuron 2: top of panel; neuron 3: bottom of
panel) in 32 trials (sorted identically for both neurons). \textbf{B)}
Average firing rate of each neuron calculated across trials in a sliding
window of length 100 ms in steps of 5 ms. \textbf{C)} Same raster plot
as in panel A with spike coincidences (i.e., pattern {[}1,1{]}) between
the two neurons marked by cyan squares. \textbf{D)} Empirical (cyan) and
expected (magenta) number of coincidences calculated in a time-resolved
manner (parameters of sliding window identical to panel B). \textbf{E)}
Time course of the surprise measure, calculated in same sliding windows
as in panel B. Surprise values that correspond to positive and negative
significance levels \(\alpha+=0.05\) and \(\alpha-=0.95\) are shown with
by horizontal red and green lines, respectively. \textbf{F)} Same raster
plot as in panel A with significant coincidences, i.e.~UEs, marked by
red squares.}\label{fig:figure2alignedPS}
\end{figure}

@@

The original publication does not provide information which event was
used as the trigger. In this experiment, 2 events that occur in every
trial could serve as trigger events, the preparatory signal PS (event
code 114) and the response signal RS (event code 124). We noticed that
the time interval between PS and RS for the longest trials was not
identical across the respective trials and varied by \(\pm\) 1 ms. Given
the UE method is applied on a time scale of 5 ms, the analysis results
therefore are expected to depend on whether trials are aligned to PS or
RS. Thus, we decide to generate the results for both alignments.

Figures~\ref{fig:figure2alignedPS} and ~\ref{fig:figure2alignedRS} show
the results of performing the UE analysis for PS- and RS-aligned data,
respectively. Here, the analysis parameters are set to the identical
values as reported in the original publication (bin size: 5 ms, analysis
time window size: 100 ms, time step of the sliding window: 5 ms,
significance level \(\alpha= 0.05\)). The comparison of the two figures
to the original figure shows agreement in the raster displays (panel A)
and the time-resolved, trial-averaged firing rate estimates (B).
However, although the graphs of the number of coincidences per sliding
window (panel D) and the surprise measure (panel E) are similar in their
overall general behavior, they differ in the details. Thus, indeed the
choice of the alignment influences the analysis result. In order to test
if one of the two alignments is in agreement with the original
publication, we perform a detailed visual comparison of our two figures
and the original one on the basis of the spikes marked as coincident
(panels C) and as part of a UE (panel D). We notice that when aligning
to PS (Figure~\ref{fig:figure2alignedPS}), the marked spikes do not
agree in all details with the original figure. However, in
Figure~\ref{fig:figure2alignedRS}, with trials aligned to RS, we find no
disagreements with the original figure. After this visual comparison we
check if our results are exactly identical to the original ones.
Therefore, we extract the positions of the data points representing
spikes and UEs in the original figure by reading the plotting commands
in the PDF file of the original paper, and compare them to our
reproduced results. In Figure~\ref{fig:validation_fig2} A the reproduced
spike times in Figure~\ref{fig:figure2alignedRS} A are plotted against
the extracted original spike times. The plotted data points lie on the
diagonal line, indicating that the reproduced spike times correspond to
to the original ones. Figure~\ref{fig:validation_fig2} B shows the same
plot for the UEs, indicating the identity of the original and reproduced
UE timings as well. To further confirm the identity, we plot in
Figure~\ref{fig:validation_fig2} C the distributions of the differences
between the original and the reproduced timings for the spikes (black)
and the UEs (red). The differences are at most +/- 0.3 ms, which are
considerably narrower than the +/- 1 ms differences caused by the
misaligned data shown in Figure~\ref{fig:figure2alignedPS} A (see for
comparison the gray plot in Figure~\ref{fig:validation_fig2} C, which
shows the differences between the original spike times and the
misaligned spike times). The remaining minor differences of the spike
and UE times of the correctly aligned data and the original data are
only due to slight errors in the extraction of the spike times from the
original figure, which is inevitable because of a limited precision of
the plotting commands in the PDF file. Thus, we confirm that spike
timings in the correctly aligned data are identical to those in the
original data, and our implementation of the UE analysis applied to
these data reproduces exactly the same results as shown in the original
paper.

\begin{figure}[htbp]
\centering
\includegraphics{validation_results_fig2.eps}
\caption{\label{fig:validation_fig2}\textbf{A)} The scatter plot of the
reproduced spike times in Figure~\ref{fig:figure2alignedRS} A plotted
against the extracted original spike times. \textbf{B)} Same plot as in
A but for the time of occurrences of UEs. \textbf{C)} The distributions
of the differences between the original and the reproduced timings for
spikes (black) and UEs (red) in Figure~\ref{fig:figure2alignedRS}. The
distribution of the differences between the original and the reproduced
timings for spikes in Figure~\ref{fig:figure2alignedPS} is shown in
gray.}\label{fig:validationux5ffig2}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{PS_aligned_matlabdata_marked.eps}
\caption{\label{fig:comparison_raster}Close-ups of the original raster
display in Figure 4A of the original publication (left) and the first
data file available at hand for reproduction (right) reveal slight
differences in the positions of some spikes. Data on the right are
aligned (similar to the data on the left) to ES1 (event code 15 in the
GDF data file). The time before the alignment event is chosen as 700 ms,
and a bin size of 5 ms is used. The red marks indicate spike clusters
with identified differences between the left and the right panel, where
at least one spike is shifted in the right panel compared to the left
one.}\label{fig:comparisonux5fraster}
\end{figure}

@@

In our reproduction of Figure 2 of the original paper we use
preprocessed data available in 1 ms resolution, that likely experienced
the \texttt{alexa2gdf} program for conversion as shown in
Figure~\ref{fig:workflow_report} (the rightmost branch), before data are
loaded into our reimplementation of the UE analysis. However, according
to the aforementioned report by MD, we only have a chance to reproduce
Figure 4A of \autocite{Riehle97} if we have the original data or a
version of them with a time resolution lower than 1 ms available. The
original raw data with 0.1 ms resolution are presumably only available
on a storage medium and format that at present we are not able to read
and interpret. However, after we contacted AR she found the data
(``jenny201\_345\_preprocessed.gdf'') of Figure 4A of
\autocite{Riehle97} with a time resolution of 0.5 ms, which likely
experienced the \texttt{2gdf} program for conversion (middle left box in
Figure~\ref{fig:workflow_report}).

We loaded this data at 0.5 ms resolution into Python and converted the
data from the 0.5 ms to the required 1 ms resolution by the mathematical
operation \(\left\lfloor x+\frac{1}{2}\right\rfloor\) , called
``rounding half up''. In numerical software packages, including IDL,
this operation is typically implemented by a function named round().
However, the round() implementation of NumPy (version 1.11.0) performs
an even rounding, i.e., values exactly halfway between two integers are
rounded to the nearest even integer. Indeed, the latter implementation
of rounding did not reproduce the result of the original publication.
Thus, we used the expression \texttt{floor(x+0.5)} to perform rounding
as it is implemented by IDL. The procedure completely reproduces panel A
of Figure 4 in the original publication (see
Figure~\ref{fig:reproducedFig4A}).

@@

We are able to reproduce the original results of \autocite{Riehle97} by
applying a new reimplementation of the Unitary Events analysis method in
Python to the original data. The method involves a number of numerical
computations and is very sensitive, as we show here by the comparison of
Figures ~\ref{fig:figure2alignedPS} and ~\ref{fig:figure2alignedRS},
which differed in the events that the trials were aligned to. This
difference in the alignment would not make any difference in the results
if the time difference between the two events (PS and RS) were identical
across the trials. But since the latter was not the case due to hardware
features of the recording setup (as we learned from the first author of
the original publication), the binning of the data started at a slightly
different time points in different trials. This likely led to a loss or
an addition of a spike in a bin and thus to a small difference of the
number of spike synchrony events (see also the discussion on the issues
of exclusive binning in \autocite{Gruen99}). In spite of this
sensitivity of the method, we succeeded in generating results that are
exactly identical to the original ones, as confirmed by the direct
comparison of the positions of spikes and UEs in the original and the
reproduced figures. This is a strong indication that our new
implementation of the analysis faithfully implemented the UE method.

@@

The reproduction of Figure 4A is more complicated than reproducing
Figure 2 due to additional reasons. First of all, the data were not
available to us. After requesting them from the first author of the
original publication we received data and were not able to reproduce the
result - in terms of the UE results, but also the data seemed slightly
different. After further consultation with the original authors we
learned that the original Figure 4A was not generated by the Matlab
implementation used for Figure 2 but by another implementation in IDL.
Both are not available to us. However, we were told that the third
author, MD, of the original publication performed a thorough comparison
of the two implementations at the time and the final report on that
investigation was made available to us. This enables us to define the
correct workflow that reproduces the original result, given we have the
data in the correct resolution at hand.

@@

\section{Author Contribution}\label{author-contribution}

\textbf{VR}: Implementing the UE analysis in Python, Reproduction of the
figures, Discussion of the results, Drafting of the manuscript, Revision
of the draft. \textbf{JI}: Reproduction of the figures (with VR),
Discussion of the results (with all other authors), Drafting of the
manuscript (with VR), Revision of the draft (with all other authors).
\textbf{MDe}: Drafting of the manuscript (with VR), Revision of the
draft (with all other authors), Assisted in integrating the UE analysis
in the Elephant library. \textbf{SG}: Conceived the original idea of
reproduction, mediated the communications with original authors,
provided suggestions for correcting errors in intermediate reproduction
results, discussed the results, revised the manuscript (with all other
authors).

\section{Acknowledgements}\label{acknowledgements}

This project received funding from EU Grant 720270 (HBP), Deutsche
Forschungsgemeinschaft Grant DE 2175/2-1 and GR 1753/4-2 of the Priority
Program (SPP 1665), the German-Japanese Computational Neuroscience
Project (German Federal Ministry for Education and Research, BMBF Grant
01GQ1114), from the Helmholtz Portfolio Theme ``Supercomputing and
Modeling for the Human Brain'', and from the Osaka Univ for the project
`Neural mechanism of active vision studied by combining large-scale
sampling of neural activity and advanced computational analysis'.
